Tensor denoising and completion based on ordinal observations

### 2020-01-28 ICML/chanwoo/ICMLdraft.tex
Add files via upload

\section{Introduction}
Multidimensional array data, a.k.a. a tensor, appears in a huge variety of applications including recommendation systems \citep{Kutty2012APR,Adomavicius2008ContextawareRS,Sun2015ProvableST},
social networks \citep{Sun2009MultiVisCS,Nickel2011ATM},
genomics \citep{Wang2017ThreewayCO},
neuroimaging (EEG, fMRI) \citep{Miwakeichi_decomposingeeg}
and signalprocessing \citep{Sidiropoulos2000ParallelFA,Cichocki2015TensorDF}.
Instead of unfolding those data tensors into matrices where many analysis methods have been proposed, we preserve its inherent multi-modal structure. Studying tensor data while respecting the structure allows us to examine complex interactions among tensor entries. Thereby we can provide extra more interpretation that cannot be addressed by traditional matrix analysis. Also, It has been shown that the tensor preserving analysis improves performance \citep{Zare2018ExtensionOP,Wang2018LearningFB}.
With those reasons, there is a growing need to develop dimension reduction method without losing multi modal structure. In the line of the attempts, a number of tensor decomposition methods have been proposed in many applications. CANDECOMP/PARAFAC (CP) decomposition was first introduced \citep{doi:10.1002/sapm19287139}
and revitalized in psychometrics \citep{Harshman1970FoundationsOT} and in linguistics \citep{Smilde2004MultiwayAW}. The Tucker decomposition was proposed in psychometrics \citep{tucker64extension,RePEc:spr:psycho:v:31:y:1966:i:3:p:279-311}.

Classical tensor completion with those decompositions has treated the entries of data as real-valued.
In many cases, However, we encounter data of which the entries are not real-valued but discrete or quantized i.e. binary-valued or multiple-valued. For example, many survey data takes the integer values. To be specific, the data in the Netflix problem has the 3 modes of `user',`movie' and `date of grade'. The entries of the data are the gradings from the users which take integer value from 1 to 5. Also, there are many cases that the data are quantized in real application. In signal processing, the data are frequently rounded or truncated so that only integer values are available. Also, in graph theory, adjacency matrix can be labeled as from 1 to 3 taking 3 when pairs of vertexes have strong connection and giving 1 when two vertexes have the weak connection according to a given threshold. If we add one more mode on adjacency matrix such as `context' or `individual', the data turns into tensor data with 3 integer values.

Therefore, performance improvement can be achieved when the observations are treated as discrete value not as continuous value.
In matrix case, there has been many achievements to complete matrix for discrete cases: Models for the case of binary or 1-bit were introduced and studied \citep{Davenport20121BitMC,Bhaskar20151bitMC}. Furthermore,  \citet{Bhaskar2016ProbabilisticLM} suggested matrix completion method for general ordinal observations.
In tensor case, however, only binary tensor has gotten an attention and achieved performance improvement using binary tensor decomposition methods \citep{Hore2016TensorDF,Wang2018LearningFB,Hong2018GeneralizedCP,Hu2018TrainingBW}. Accordingly, a general method for the data which has more than 2 ordered label is needed.

We organized this paper as follows. In section 2, we discuss detailed assumptions and descriptions about our probabilistic model in \eqref{multinomial}. Also, we suggest the estimation method for the latent parameters and related algorithm. In section 3, we provide the statistical properties of the upper, lower bounds and the phase-transition. We then provide the numerical experiments. Our model is applied to  real-world data to check validity and performance in section 5. Finally, we wrap up the paper with a discussion.

@@

\section{Model}
\subsection{Low-rank cumulative model}
For the K-mode ordinal tensor $\mathcal{Y} = [\![ y_\omega]\!]\in[L]^{d_1\times\cdots\times d_K}$, we assume that its entries are realizations of independent multinomial random variables, such that
\begin{equation}
    \begin{aligned}
        \label{mainmodel}
        \mathbb{P}(y_\omega=l) = f_l(\theta_\omega), \quad \omega\in [d_1]\times\cdots\times[d_K].
    \end{aligned}
\end{equation}
In this model, a twice differentiable function $f_l : \mathbb{R}\rightarrow [0,1]$ with $l\in [L]$ is strictly increasing and satisfies $\sum_l f_l(\theta) = 1$ for a fixed $\theta$.
The tensor $\Theta = [\![\theta_\omega]\!]\in \mathbb{R}^{d_1\times\cdots\times d_K}$
is a hidden parameter which we are interested in. We assume the parameter tensor $\Theta$ is continuous value and admits a rank $\mr = (r_1,\cdots,r_K)$ Tucker decomposition,
\begin{equation}
    \begin{aligned}
        \label{tucker}
        \Theta = \mathcal{C}\times_1\mM_1\cdots\times_K\mM_N,
    \end{aligned}
\end{equation}
where $\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_K}$ is a core tensor and $\mM_k\in \mathbb{R}^{d_k\times r_k}$, for $n \in [L]$ is a factor matrix. We can get information about the influence of each mode by checking factor matrices.

The tensor model \eqref{mainmodel} has an equivalent representation as a latent model with K-level quantization. \citep{Davenport20121BitMC,Lan2014MatrixRF,Bhaskar20151bitMC,Cai2013AMC} where
$\mathcal{Y} = [\![ y_\omega]\!]$ is a quantized value such that,
\begin{equation}
    \label{latentmd}
\begin{aligned}
    y_\omega = \mathcal{Q}(\theta_\omega+\epsilon_\omega),\quad \omega\in [d_1]\times\cdots[d_K],
\end{aligned}
\end{equation}
where $\mathcal{E} = [\![\epsilon_\omega]\!]$ is a noise tensor of $i.i.d.$ from cumulative distribution function $\Phi(\cdot)$ and the function $\mathcal{Q}:\mathbb{R}\rightarrow [L]$ is a quantizer having the following rule.
\begin{equation}
    \label{qfunction}
    \begin{aligned}
        \mathcal{Q}(x) = l,\quad \text{ if } b_{l-1}< x \leq b_{l},\quad l\in [L],
    \end{aligned}
\end{equation}
where $\mb = (b_1,\cdots,b_{L-1})$ is a cutpoints vector such that $ -\infty = b_0<b_1<\cdots<b_L = \infty$. That is, the entries of observed tensor $\mathcal{Y}$ fall in category $l$ when the associated entries of the latent tensor $\Theta+\mathcal{E}$ fall in the $l$th interval of values. Then we have
\begin{equation}
    \label{cumulative}
    \begin{aligned}
        f_l(\theta_\omega) &= \mathbb{P}(y_\omega = l)\\
        &= \Phi(b_l-\theta_\omega)-\Phi(b_{l-1}-\theta_\omega).
    \end{aligned}
\end{equation}
We can diversify our model by the choices of $\Phi(\cdot)$, or equivalently the distribution of $\mathcal{E}$ with given $\mb$.
The followings are 2 common choices of $\Phi(\cdot)$.
\begin{example}[Logistic link/Logistic noise] The logistic model is represented by \eqref{mainmodel} with $f_l(\theta) = \Phi_{log}(\frac{b_l-\theta}{\sigma})-\Phi_{log}(\frac{b_{l-1}-\theta}{\sigma})$  where $\Phi_{log}(x/\sigma) = (1+e^{-x/\sigma})$. Equivalently, the noise $\epsilon_\omega$ in \eqref{latentmd} follows i.i.d. logistic distribution with the scale parameter $\sigma.$
\end{example}

\begin{example}[Probit link/Gaussian noise] The probit model is represented by \eqref{mainmodel} with $f_l(\theta) = \Phi_{norm}(\frac{b_l-\theta}{\sigma})-\Phi_{norm}(\frac{b_{l-1}-\theta}{\sigma})$ where $\Phi_{norm}$ is the cumulative distribution function of the standard Gaussian. Equivalently, the noise $\epsilon_\omega$ in \eqref{latentmd} follows i.i.d.$N(0,\sigma^2)$.
\end{example}

\subsection{Rank-constrained likelihood-based estimation}
Our goal is to estimate unknown parameter tensor $\Theta$ and a cutpoints vector $\mb$ from observed tensor $\mathcal{Y}$ using a constrained likelihood approach. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega = [d_1]\times\cdots\times[d_K]$ or a random subset indeuced from the subsampling distribution. The log-likelihood function for \eqref{mainmodel} is
\begin{equation}
    \label{likelihood}
    \begin{aligned}
    \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb) = \sum_{\omega\in\Omega}\Big[\sum_{l\in[L]}\log(f_l(\theta_\omega))\mathds{1}_{\{y_\omega = l\}}\Big],
    \end{aligned}
\end{equation}
where the cutpoints vector $\mb$ is implicitly contained in the function $f_l$.
Considering the Tucker structure in \eqref{tucker}, we have the following constrained optimization problem.
\begin{equation}
    \label{eq:optimization}
    \begin{aligned}
        &\max_{(\Theta, \mb)\in \mathcal{D}\times\mathcal{B}}\mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb),\text{ where }\\
        &\mathcal{D} = \{\Theta\in \mathbb{R}^{d_1\times\cdots\times d_K}: \rank(\Theta) = \bm{r} \text{ and } \mnormSize{}{\Theta}\leq \alpha \}\\
        &\mathcal{B} = \{\mb\in\mathbb{R}^{L-1}:  b_1<\cdots < b_{L-1}\},
    \end{aligned}
\end{equation}
for a given rank $\bm r\in \mathbb{N_+}^K$ and a bound $\alpha \in \mathbb{R}_+$. The search space $\mathcal{D}$ has two constraints on unknown parameter $\Theta$. The first constraint ensures that the unknown parameter $\Theta$ admits the Tucker decomposition with rank $\bm r$. The second constraint makes the entries of $\Theta$ bounded by a constant $\alpha$. This bound condition is a technical assumption to help to recover $\Theta$ in the noiseless case. Similar conditions has been imposed in many literatures for the matrix case \citep{Davenport20121BitMC,Bhaskar20151bitMC,Cai2013AMC,Bhaskar2016ProbabilisticLM}.
The search space $\mathcal{B}$ makes sure that the probability function $f_l$ in \eqref{cumulative} is positive.

\subsection{Optimization}
In this section, we describe the algorithm to seek the optimizer of \eqref{eq:optimization}. The objective function $\mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb)$ is concave in $(\Theta,\mb)$ whenever $\Phi(x)$ is log-concave \citep{McCullagh1980RegressionMF,Burridge1981ANO}.
However, the feasible set $\mathcal{D}$ is not a convex set, which makes the optimization \eqref{eq:optimization} a non-convex problem. One approach to handle this problem is utilizing the Tucker decomposition and converting optimization into a block-wise convex problem.
Let us denote the objective function as
\begin{equation}
    \label{blockwise}
    \begin{aligned}
         \mathcal{L}(\mathcal{C},\mM_1,\cdots,\mM_K,\mb) = \mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C}\times_1\mM_1\cdots\times_K\mM_K,\mb).
    \end{aligned}
\end{equation}
From \eqref{blockwise}, we have $K+2$ blocks of variables in the objective function, one for the cutpoints vector $\mb$, one for the core tensor $\mathcal{C}$ and  K for the factor matrices $\mM_K$'s.
We can change the optimization problem to simple convex problem if any $K+1$ out of the $K+2$ blocks being fixed. Therefore, we can alternatively update one block at a time while other blocks being fixed.
The algorithm \ref{alg} gives the full description.

\begin{algorithm}[tb]
        \caption{Ordinal tensor decomposition }\label{alg}
        \begin{algorithmic}[]
            \STATE{\bfseries Input:}  \text{ Ordinal tensor }
            $\mathcal{Y}\in [L]^{d_1\times\cdots\times d_K}
            $,\\ \hspace{.43in} Rank $\mr\in \mathbb{N_+}^{K-1}$,\\
            \hspace{.43in} Entry-wise bound $\alpha\in \mathbb{R_+}$.
            \STATE{\bfseries Output:} $(\hat\Theta,\hat{\mb}) =  \argmax_{(\Theta,\mb)\in \mathcal{D}\times\mathcal{B}}  \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb).$
            \STATE Initialize  Core tensor $\mathcal{C}^{(0)}$,\\
            \hspace{.5in} Factor matrices $ \{\mM_1^{(0)},\cdots,\mM_K^{(0)}\}$,\\
            \hspace{.5in} Cut-off points $\mb^{(0)}$.
            \FOR{t = 1,2,$\cdots$,}
            \FOR{k = 1,2,$\cdots$,K}
            \STATE{Update $\mM_k$ fixing other blocks.}
            \STATE $\mM_k^{(t+1)}\gets\argmax_{\mM_k\in\mathbb{R}^{d_k\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mM_k)$ s.t.
             $\|\mathcal{C}^{(t)}\times_1\mM_1^{(t+1)}\cdots\times_k\mM_k\cdots\times_K\mM_K^{(t)}\|_\infty \leq \alpha$
            \ENDFOR
            \STATE {Update $\mathcal{C}$ fixing other blocks.}
            \STATE $\mathcal{C}^{(t+1)} \gets \argmax_{\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C})$
            s.t.\newline $\|\mathcal{C}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}\|_\infty \leq \alpha$

            \STATE $\Theta^{(t+1)} \gets \mathcal{C}^{(t+1)}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}$
            \STATE {Update $\mb$ fixing $\Theta^{(t+1)}$.}
            \STATE $\mb^{(t+1)} \gets \argmax_{\mb\in \mathbb{R}^{L-1}}\mathcal{L}_{\mathcal{Y},\Omega}\big(\Theta^{(t+1)},\mb\big)$
            \ENDFOR
            \STATE \textbf{return}
            $\Theta,\mb$
        \end{algorithmic}
    \end{algorithm}

\subsection{Rank selection}
Algorithm \ref{alg} takes  the rank $\bm r$ as an input variable. In practice, the rank $\bm r$ is hardly known. Estimating an appropriate rank $\bm r$ from a given tensor is an important issue. We suggest to use Bayesian Information Criterion(BIC) to choose the rank.
\begin{equation}
    \label{bic}
    \begin{aligned}
        \hat{\bm r} &= \argmin_{\bm r\in \mathbb{N}_+^K} BIC(\bm r)\\
        &=\argmin_{\bm r\in \mathbb{N}_+^K}\bigg[-2\mathcal{L}_\mathcal{Y}(\hat\Theta(\bm r)) \quad+
        \\&\qquad\big(\prod_{k\in[K]} r_k+\sum_{k\in[K]}(d_kr_k-r_k^2)\big)\log(\prod_{k\in[K]} d_k)\bigg],
    \end{aligned}
\end{equation}
where $\hat\Theta(\bm r)$ is a maximum likelihood estimater given the rank $\bm r$. The second part of the summation in \eqref{bic} is the number of  independent parameters in the model. We select the rank $\hat{\bm r}$ that minimizes BIC value through the grid search method.
\section{Real-world Data Applications}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connectome Project (HCP) \citep{Geddes2016HumanBM}. In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts \citep{Baltrunas2011InCarMusicCM}.
\subsection{Human Connectome Project (HCP)}
The human connectome project (HCP) is a $68\times 68 \times 136$ tensor where the first two modes have 68 indices representing brain regions and the last mode has 136 indices meaning individuals. All the individual images were preprocessed following a standard pipeline \citep{Zhang2018MappingPS}, and the brain was parcellated to 68 regions of interest following the Desikan atlas \citep{Desikan2006AnAL}. The tensor entries consist of $\{1,2,3\}$, the strength of fiber connections between 68 brain regions for each of the 136 individuals. First, we apply our ordinal tensor decomposition method with a logistic link function to the HCP data and identify similar brain nodes based on latent parameters. The BIC result suggests $\bm r = (23,23,8)$  with $\mathcal{L}_\mathcal{Y}(\hat{\Theta},\hat{\mb}) = -216645.8$.

Also, we compare our ordinal tensor decomposition method with a logistic link function with other methods: the continuous tensor decomposition method, the 1 bit-tensor completion method with probit link function and the 1 bit-tensor completion method with logistic link function. We use 5 folded cross validation method for the comparison. Specifically, we randomly split the tensor entries into 5 similar sized pieces and alternatively use each piece of entries as a test data using other $80\%$ entries as a training data. The entries in the test data are encoded as missing and then predicted based on each method from the training data. We set the rank of the tensor as $\bm r = (23,23,8)$ for the tensor decomposition methods which minimizes BIC value. Table \ref{CV} shows RMSE, MAE and error(false prediction rate), averaged over 5 test set results. We can check that our ordinal tensor decomposition model outperforms the other methods in all criteria.

>>>>

\section{Introduction}
Multidimensional array datasets, a.k.a. a tensor, appear in a huge variety of applications including recommendation systems \citep{kutty2012people,adomavicius2011context,sun2017provable},
social networks \citep{sun2009multivis,nickel2011three},
genomics \citep{wang2018learning},
neuroimaging (EEG, fMRI) \citep{miwakeichi2004decomposing}
and signalprocessing \citep{sidiropoulos2000parallel,cichocki2015tensor}.
Instead of unfolding those data tensors into matrices where many analysis methods have been proposed, we suggest preserving multi-modal tensor structure in this paper. Studying tensor data while respecting the structure allows us to examine complex interactions among tensor entries. Thereby we can expect to provide extra more interpretation that cannot be addressed by traditional matrix analysis. Furthermore, It has been shown that the tensor preserving analysis improves performance \citep{zare2018extension,wang2018learning}.
With those reasons, there is a growing need to develop dimension reduction method without losing tensor structure. In the line of the attempts, a number of tensor decomposition methods have been proposed in many applications. CANDECOMP/PARAFAC (CP) decomposition was first introduced \citep{hitchcock1928multiple}
and made use of in psychometrics \citep{harshman1970foundations} and in linguistics \citep{smilde2005multi}. The Tucker decomposition was proposed in psychometrics \citep{tucker1964extension,tucker1966some}.

Classical tensor completion with those decompositions has treated the entries of data as real-valued.
In many cases, However, we encounter data sets of which the entries are not real-valued but discrete or quantized i.e. binary-valued or ordinal-valued. For example, many survey data sets take the integer values. To be specific, the data set in the Netflix problem has the three modes of `user',`movie' and `date of grade'. The entries of the data are the ratings from the users which take integer value from 1 to 5. Another example is the case where the data sets are quantized in real application. In signal processing, the data sets are frequently rounded or truncated so that only integer values are available. In network analysis, entries of adjacency matrix can be labeled as from 1 to 3 taking 3 when pairs of vertexes have strong connection and giving 1 when two vertexes have the weak connection according to a given threshold. If we add one more mode on adjacency matrix such as `context' or `individual', the data turns into tensor data with 3 integer values.

We expect that performance improvement can be achieved when the observations are treated as discrete ordinal value not a continuous value.
In matrix case, there has been many achievements to complete matrix for discrete cases: Models for the case of binary or 1-bit were introduced and studied \citep{davenport20141,bhaskar20151}. Furthermore,  \citet{bhaskar2016probabilistic} suggested matrix completion method for general ordinal observations.
In tensor case, however, only binary tensor has gained an attention and achieved performance improvement using binary tensor decomposition methods \citep{hore2016tensor,wang2018learning,hong2018generalized,hu2018training}. Accordingly, a general method for the data which has more than 2 ordered label is needed.

We organized this paper as follows. We specify our notations that we will use over this paper in the next section.
In section 3, we discuss detailed assumptions and descriptions about our probabilistic model. Also, we suggest the estimation method for the latent parameters and related algorithm.
In section 4, we provide the statistical properties of the upper, lower bounds and the phase-transition.
We then provide the numerical experiments.
Our model is applied to  real-world data to check validity and performance in section 6.
Finally, we wrap up the paper with a discussion.

@@

\section{Model}
\subsection{Low-rank ordinal tensor model}
For the $K$-mode ordinal tensor $\mathcal{Y} = [\![ y_\omega]\!]\in[L]^{d_1\times\cdots\times d_K}$, we assume that its entries are realizations of independent multinomial random variables, such that
\begin{equation}
    \begin{aligned}
        \label{mainmodel}
        \mathbb{P}(y_\omega\leq \ell) = f(\theta_\omega+b_\ell), \quad \omega\in [d_1]\times\cdots\times[d_K].
    \end{aligned}
\end{equation}
In this model, a twice differentiable function $f: \mathbb{R}\rightarrow [0,1]$  is known and a cumulative distribution function and $\mb = (b_1,\cdots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots<b_{L-1}$. We refer to $\mb$ as the cut-off points and $f$ the link function.
The tensor $\Theta = [\![\theta_\omega]\!]\in \mathbb{R}^{d_1\times\cdots\times d_K}$
is a hidden parameter which we are interested in. We assume that the entries of the parameter tensor $\Theta$ are continuous valued and $\Theta$ admits a rank $\mr = (r_1,\cdots,r_K)$ Tucker decomposition,
\begin{equation}
    \begin{aligned}
        \label{tucker}
        \Theta = \mathcal{C}\times_1\mM_1\cdots\times_K\mM_N,
    \end{aligned}
\end{equation}
where $\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_K}$ is a full rank core tensor and $\mM_k\in \mathbb{R}^{d_k\times r_k}$, for $n \in [L]$ is a factor matrix with orthogonal columns. This low rank tensor structure allows us to reduce dimension and complete tensors from available entries.

The ordinal tensor model \eqref{mainmodel} has the equivalent representation as a latent model with $L$-level quantization. \citep{davenport20141,lan2014matrix,bhaskar20151,cai2013max} where the entry of
$\mathcal{Y} = [\![ y_\omega]\!]$ is a quantized value such that,
\begin{equation}
    \label{latentmd}
\begin{aligned}
    y_\omega = \mathcal{Q}(\theta_\omega+\epsilon_\omega),\quad \omega\in [d_1]\times\cdots[d_K],
\end{aligned}
\end{equation}
where $\mathcal{E} = [\![\epsilon_\omega]\!]$ is a noise tensor i.i.d. entries from cumulative distribution function $\mathcal{F}_\epsilon$ such that $f(\theta) = \mathbb{P}(\epsilon \geq -\theta)$ and  $\mathcal{Q}:\mathbb{R}\rightarrow [L]$ is a quantizer having the following rule.
\begin{equation}
    \label{qfunction}
    \begin{aligned}
        \mathcal{Q}(x) = \ell,\quad \text{ if } b_{\ell-1}< x \leq b_{\ell},\quad \text{ for all } \ell\in [L],
    \end{aligned}
\end{equation}
That is, the entries of observed tensor $\mathcal{Y}$ fall into category $\ell$ when the associated entries of the latent tensor $\Theta+\mathcal{E}$ fall into the $\ell$-th interval of values. From this point of view, we can see the latent parameter $\Theta$ as the signal prior to contamination and quantization.
We can diversify our model by the choices of $f$, or equivalently the distribution of $\mathcal{E}$.
The followings are two common choices of $f$.

\begin{example}[Logistic link/Logistic noise] The logistic model is represented by \eqref{mainmodel} with $f(\theta) = \Phi_{log}(\theta/\sigma)$  where $\Phi_{log}(x/\sigma) = (1+e^{-x/\sigma})$. Equivalently, the noise $\epsilon_\omega$ in \eqref{latentmd} follows i.i.d. logistic distribution with the scale parameter $\sigma.$
\end{example}

\begin{example}[Probit link/Gaussian noise] The probit model is represented by \eqref{mainmodel} with $f(\theta) = \Phi_{norm}(\theta/\sigma)
$ where $\Phi_{norm}$ is the cumulative distribution function of the standard Gaussian. Equivalently, the noise $\epsilon_\omega$ in \eqref{latentmd} follows i.i.d.$N(0,\sigma^2)$.
\end{example}

\subsection{Rank-constrained likelihood-based estimation}
Our goal is to estimate unknown parameter tensor $\Theta$ and cut-off points $\mb$ from observed tensor $\mathcal{Y}$ using a constrained likelihood approach. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega = [d_1]\times\cdots\times[d_K]$ or a random subset induced from the subsampling distribution. The log-likelihood function for \eqref{mainmodel} is
\begin{equation}
    \label{likelihood}
    \begin{aligned}
    \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb) = \sum_{\omega\in\Omega}\Big[\sum_{\ell\in[L]}\log(g_\ell(\theta_\omega))\mathds{1}_{\{y_\omega = \ell\}}\Big],
    \end{aligned}
\end{equation}
where $g_\ell(x) = \mathbb{P}(y_\omega = \ell) = f(\theta_\omega+b_\ell)-f(\theta_\omega+b_{\ell-1})$ from \eqref{mainmodel}. We define $b_0 = -\infty,b_L = \infty$ so that $f(b_0) = 0, f(b_L) = 1$. The cut-off points $\mb$ is implicitly contained in the function $g_\ell$.
Considering the Tucker structure in \eqref{tucker}, we have the following constrained optimization problem.
\begin{equation}
    \label{eq:optimization}
    \begin{aligned}
        &\max_{(\Theta, \mb)\in \mathcal{D}\times\mathcal{B}}\mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb),\text{ where }\\
        &\mathcal{D} = \{\Theta\in \mathbb{R}^{d_1\times\cdots\times d_K}: \rank(\Theta) \leq \bm{r} \text{ and } \mnormSize{}{\Theta}\leq \alpha \}\\
        &\mathcal{B} = \{\mb\in\mathbb{R}^{L-1}:  b_1<\cdots < b_{L-1}\},
    \end{aligned}
\end{equation}
for a given rank $\bm r\in \mathbb{N_+}^K$ and a bound $\alpha \in \mathbb{R}_+$. The search space $\mathcal{D}$ has two constraints on unknown parameter $\Theta$. The first constraint ensures that the unknown parameter $\Theta$ admits the Tucker decomposition with rank $\bm r$. The second constraint makes the entries of $\Theta$ bounded by a constant $\alpha$. This bound condition is a technical assumption to help to recover $\Theta$ in the noiseless case. Similar conditions has been imposed in many literatures for the matrix case \citep{davenport20141,bhaskar20151,cai2013max,bhaskar2016probabilistic}.
The search space $\mathcal{B}$ makes sure that the probability function $g_\ell$ in \eqref{likelihood} is strictly positive.

\subsection{Optimization}
In this section, we describe the algorithm to seek the optimizer of \eqref{eq:optimization}. The objective function $\mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb)$ is concave in $(\Theta,\mb)$ whenever $f(x)$ is log-concave \citep{mccullagh1980regression,burridge1981note}.
However, the feasible set $\mathcal{D}$ is not a convex set, which makes the optimization \eqref{eq:optimization} a non-convex problem. One approach to handle this problem is utilizing the Tucker decomposition and converting optimization into a block-wise convex problem.
From \eqref{likelihood} and \eqref{tucker}, we have $K+2$ blocks of variables in the objective function, one for the cut-points vector $\mb$, one for the core tensor $\mathcal{C}$ and  $K$ for the factor matrices $\mM_K$'s.
We can change the optimization problem to simple convex problem if any $K+1$ out of the $K+2$ blocks being fixed. Therefore, we can alternately update one block at a time while other blocks being fixed.
The algorithm \ref{alg} gives the full description.

\begin{algorithm}[]
        \caption{Ordinal tensor decomposition }\label{alg}
        \begin{algorithmic}[]
            \STATE{\bfseries Input:}  \text{ Ordinal tensor }
            $\mathcal{Y}\in [L]^{d_1\times\cdots\times d_K}
            $,\\ \hspace{.43in} Rank $\mr\in \mathbb{N_+}^{K-1}$,\\
            \hspace{.43in} Entry-wise bound $\alpha\in \mathbb{R_+}$.
            \STATE{\bfseries Output:} $(\hat\Theta,\hat{\mb}) =  \argmax_{(\Theta,\mb)\in \mathcal{D}\times\mathcal{B}}  \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb).$
            \STATE Initialize  Core tensor $\mathcal{C}^{(0)}$,\\
            \hspace{.5in} Factor matrices $ \{\mM_1^{(0)},\cdots,\mM_K^{(0)}\}$,\\
            \hspace{.5in} Cut-off points $\mb^{(0)}$.
            \FOR{t = 1,2,$\cdots$,}
            \FOR{k = 1,2,$\cdots$,K}
            \STATE{Update $\mM_k$ while fixing other blocks:}
            \STATE $\mM_k^{(t+1)}\gets\argmax_{\mM_k\in\mathbb{R}^{d_k\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mM_k)$
            \newline
             s.t. $\mnormSize{}{\Theta^{(t+1)}}\leq \alpha$, where $\Theta^{(t+1)}$ is the parameter tensor based on the current block estimates.
            \ENDFOR
            \STATE {Update $\mathcal{C}$ while fixing other blocks:}
            \STATE $\mathcal{C}^{(t+1)} \gets \argmax_{\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C})$
            s.t.  $\mnormSize{}{\Theta^{(t+1)}}\leq \alpha$
            \STATE Update $\Theta$ based on the current block estimates:
            \STATE $\Theta^{(t+1)} \gets \mathcal{C}^{(t+1)}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}$
            \STATE {Update $\mb$ while fixing $\Theta^{(t+1)}$:}
            \STATE $\mb^{(t+1)} \gets \argmax_{\mb\in \mathbb{R}^{L-1}}\mathcal{L}_{\mathcal{Y},\Omega}\big(\mb\big)$
            \ENDFOR
            \STATE \textbf{return}
            $\Theta,\mb$
        \end{algorithmic}
    \end{algorithm}

\subsection{Rank selection}
Algorithm \ref{alg} takes  the rank $\bm r$ as an input variable. In practice, the rank $\bm r$ is hardly known. Estimating an appropriate rank $\bm r$ from a given tensor is an important issue. We suggest to use Bayesian Information Criterion(BIC) to choose the rank.
\begin{equation}
    \label{bic}
    \begin{aligned}
        \hat{\bm r} &= \argmin_{\bm r\in \mathbb{N}_+^K} BIC(\bm r)\\
        &=\argmin_{\bm r\in \mathbb{N}_+^K}\bigg[-2\mathcal{L}_{\mathcal{Y},\Omega}(\hat\Theta(\bm r),\hat\mb(\mr)) +
        p_e(\mr)\log\big(\prod_{k\in[K]} d_k\big)\bigg],
    \end{aligned}
\end{equation}
where $\hat\Theta(\bm r),\hat\mb(\mr)$ is a maximum likelihood estimate given the rank $\bm r$, and $p_e(\rm)\eqdef \sum_k(d_k-r_k)r_k = \prod_k r_k$ is the effective number of free parameters in the model. We select the rank $\hat{\bm r}$ that minimizes BIC value through the grid search method.
\section{Real-world Data Applications}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connectome Project (HCP) \citep{geddes2016human}. In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts \citep{baltrunas2011incarmusic}.
\subsection{Human Connectome Project (HCP)}
The human connectome project (HCP) is a $68\times 68 \times 136$ tensor where the first two modes have 68 indices representing brain regions and the last mode has 136 indices meaning individuals. All the individual images were preprocessed following a standard pipeline \citep{zhang2018mapping}, and the brain was parcellated to 68 regions of interest following the Desikan atlas \citep{desikan2006automated}. The tensor entries consist of $\{1,2,3\}$, the strength of fiber connections between 68 brain regions for each of the 136 individuals. We apply our ordinal tensor decomposition method with a logistic link function to the HCP data. The BIC result suggests $\bm r = (23,23,8)$  with $\mathcal{L}_\mathcal{Y}(\hat{\Theta},\hat{\mb}) = -216645.8$.

We check the performance of our model and compare with other methods: the continuous tensor decomposition method, the 1 bit-tensor completion method with probit link function and the 1 bit-tensor completion method with logistic link function. We use 5 folded cross validation method for the comparison. Specifically, we randomly split the tensor entries into 5 similar sized pieces and alternately use each piece of entries as a test data using other $80\%$ entries as a training data. The entries in the test data are encoded as missing and then predicted based on each method from the training data. In our model,  predict $y_\omega$ as $\hat{y}_\omega^{\text{(mode)}} = \argmax_\ell(g_\ell(
\hat\theta_\omega,\hat\mb)).$ Table \ref{CV} shows RMSE, MAE and error(false prediction rate), averaged over 5 test set results. We can check that our ordinal tensor decomposition model outperforms the other methods in all criteria.

\begin{table}[t]
\label{CV}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
 Method & MSE & MAD & MCR \\
\midrule
\begin{tabular}[c]{@{}c@{}}Oridnal tensor\\ Decomposition\end{tabular}  & 0.1503 & 0.1502& 0.1502\\ \\
\begin{tabular}[c]{@{}c@{}}Conti tensor\\ Decomposition\end{tabular} & 0.1603& 0.1600& 0.1598\\ \\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (probit)\end{tabular}  & 0.3658 & 0.3632 & 0.3619\\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (logistic)\end{tabular}  &  0.4514 & 0.4215 & 0.4066
\\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\caption{Results of comparisons among 4 methods on the HCP data predicting the test data. Four methods are the ordinal tensor decomposition algorithm, the continuous tensor decomposition algorithm and the 1 bit tensor completion method with probit link function and logistic link function. Each method is evaluated by RMSE, MAE, error(false prediction rate).}
\end{table}
